Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-07-29 extended_catalog/paper/extended_catalog.tex
improvements
adding stefan's introduciton

Using two years of all sky surveying data, the LAT Collaboration
published a catalog of 1873 \gev sources.
Many of these were associated
with a variety of source classes including Active Galactic Nuclei (AGN),
pulsars, galaxies, and supernova remnants.  Many of these classes can be
spatially resolved when observed at other frequencies and could possibly
be spatially resolvable in \gev by the LAT.

The LAT collaboration has previously reported on the discovered of
five spatially extended supernova remnants (SNRs) IC443, W28, W44,
W51C, and RX J1713.7-2946 (\cite{ic443,w28,w44,w51c,rx_j1713_lat}).
The LAT Collaboration has reported on the detection of three extended
pulsar wind nebulae (PWNe) MSH 15-52 and Vela X, and HESS J1825-137
(\cite{msh1552,velax,fermi_hess_j1825}), the Large Magellanic
and Small Magalenic Cloud (\cite{lmc,smc}), and the radio galaxy
Centarus A (\cite{cen_a_lat}) .  There are other source classes which
could possibly be spatially resolved by the LAT including galaxy clusters
and dark matter satellite galaxies.

Furthermore, a variety of spatially extended but otherwise unassociated
sources have been detected in the Galactic plane at \tev energies using
Air Cherenkov detectors. Most prominent was a survey of the Galactic
plane using the High Energy Stereoscopic System (H.E.S.S) which presented
14 spatially extended sources (\cite{hess_plane_survey}) sources with
extensions varying from $~0.1\deg$ to $~0.25\deg$. Most \tev sources
in the galaxy are spatially extended and it is expected that these same
sources would have have extended \gev emission.

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large point spread
function (PSF), there are often several possible source counterparts.
Finding a matching morphology can be used to uniquely identify the source.
Because of the largely varying PSF of the LAT, the spatial and spectral
information about a source do not nicely decouple. Therefore, a biased
spatial model will bias the spectral model of the source and studying
the extension of the source can correct for this.  Furthermore, correctly
modeling a source's extension is important for improving the overall model
of the sky and removing excess residuals.  These residuals manifested
themselves in 1FGL as spurious point sources in the region region
surrounding the Large Magellanic Cloud (\cite{first_cat}).  For this
reason, we are motivated to take the 1873 sources that were detected in
2FGL and apply a search for those sources which are spatially extended.

@@

\todo[inline]{note about how MSH 15-52 can't go lower in
energy because of nearby pulsar.}

@@

This source is spatially coincident with the extended \tev source HESS
J1616-508 (\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1613.6-5100c},
contours of HESS J1616-508 are overlaid on 2FGL J1615.0-5051.
The H.E.S.S. experiment measured an extension $0.136\pm 0.008$
when fitting the source with an elliptical Gaussian spatial model.
This size corresponds to a 68\% source containment radius of
$\rsixeight\sim0.21\deg\pm0.01$. This size is comparable to the LAT size
$\rsixeight=0.27\deg\pm0.03$ (see section~\ref{compare_source_size}).

@@

This source is spatially coincident with the extended \tev source HESS
J1614-518 (\cite{hess_plane_survey}). In figure~\ref{1FGL_J1613.6-5100c},
contours of HESS J1614-518 are overlaid on 2FGL J1615.2-5138.
The H.E.S.S. experiment measured a Gaussian extension of
$\sigma=0.23\deg\pm0.02\deg$ and $\sigma=0.15\pm0.02$ in the semi-major
and semi-minor axis. This size corresponds to a 68\% containment size
of $\rsixeight=0.35\pm0.03\deg$ and $0.23\pm0.03$.  This elliptical size
matches well the LAT size of $\rsixeight\sim0.34\pm0.02$.

@@

This extended source is spatially coincident with the extended \tev source HESS
J1632-478 (\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1632.9-4802c},
contours of  HESS J1632-478 are overlaid on 2FGL J1635.4-4717c.
H.E.S.S measured an extension of $\sigma=0.21\pm0.05$ and
$0.06\pm0.04$ along the semi-major and semi-minor axes when fitting the source
with an elliptical Gaussian spatial model.  This corresponds to a 68\%
containment size of $\rsixeight=0.31\pm0.08$ and $0.09\pm0.06$ along the
semi-major and semi-minor axis. This size is consistent with the LAT size of
$\rsixeight=0.36\pm0.03$.

@@

This source is spatially coincident with the \tev source HESS J1837-069
(\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1837.5-0659c},
contours of are H.E.S.S J1837-069 are overlaid on 2FGL
J1837.3-0700c. H.E.S.S. measured an extension of $\sigma=0.12\pm0.02$
and $0.05\pm0.02$ along the semi-major and semi-minor axis when fitting
the source with an elliptical Gaussian spatial model.  This corresponds to
a 68\% containment radius of $\rsixeight=0.18\pm0.03$ and $0.08\pm0.03$
along the semi-major and semi-minor axis. The size is comparable to
LAT which fit a 68\% containment radius of $\rsixeight=0.29\pm0.07$.
The LAT size is larger than the H.E.S.S. size but the sizes are almost
almost within the $1\sigma$ errors. It is possible that the morphologies
are different but this is not very significant.  A possible reason is
confusion could be that the LAT source may be partially fitting other
sources in the region.

@@

This source is spatially conincident with the SNR RX J1713.7-2946 and was
recently reported by (\cite{rx_j1713_lat}).  To analyze this source,
we used the same background model as the recent LAT publication.
2FGL J1715.4-4024c is spatially coincident with Source A and was
moved to the published location of $(ra,dec)=(258.84,-40.46)$. Source
B was added at $(ra,dec)=(258.71,-38.70)$ and Source C was added at
$(ra,dec)=(257.47,-39.75)$.  Figure \ref{2FGL_J1712.4-3941} shows
the position of these nearby sources on top of a contour of SNR RX
J1713.7-2946 seen in \tev by H.E.S.S (\cite{rx_j1713_hess}).

@@

\begin{figure}
  \begin{center}
    \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_RX_J1713.7-3946}
    % this plot came from
  % /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/source_plots/RX_J1713.7-3946/v6/run.py
  \end{center}
  \caption{A Galactic and isotropic diffuse subtracted 1 \gev to 100
  \gev counts map of the region around 2FGL J1712.4-3941 smoothed by a
  0.25\deg Gaussian kernel.  This source is spatially coincident with
  RX J1713.7-2946 was was recently reported by (\cite{rx_j1713_lat}).
  The light blue contours are the \tev emission reported by
  H.E.S.S. (\cite{rx_j1713_hess}).  The region was analyzed with the
  same background model as (\cite{rx_j1713_lat}).  Source A (blue cross)
  is spatially coincident with 2FGL J1715.4-4024c (blue star) so 2FGL
  J1715.4-4024c was moved to the position of source A.  The grey crosses
  represent from left to right the position of source B and C which were
  added to the background model. Plot (a) has the diffuse emisission model
  subtracted and plot (b) has the diffuse emission and all background
  sources also subtracted.
  }\label{2FGL_J1712.4-3941}
\end{figure}

@@

@@

\clearpage
\begin{figure}
  \begin{center}
    \includegraphics{summary_plots/gev_vs_tev_plot.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_tev_gev_sizes/v1/run.py
    \end{center}
    \caption{
    A comparison of the \gev and \tev sizes of LAT
    extended sources detected by \tev instruments.  The \tev
    extensions of W30, 2FGL J1837.3-0700c, 2FGL J1632.4-4753c,
    2FGL J1615.0-5051, and 2FGL J1615.2-5138 are from the
    H.E.S.S. Galactic plane survey (\cite{hess_plane_survey}).
    The \tev extensions of MSH 15-52, HESS J1825-137, Vela X,
    Vela Jr., RX J1713.7-2946 and W28 are respectively from H.E.S.S.
    (\cite{msh_15_52_hess,hess_j1825_hess,vela_x_hess,vela_jr_hess,rx_j1713_hess,w28_hess}).
    The \tev extension of IC443 is from Veritas (\cite{ic443_veritas})
    and W51c is from Magic (\cite{w51c_with_magic_at_fermi_symposium}).
    The LAT extension of Vela X is from (\cite{velax}).  Except for RX
    J1713.7-2946 Vela Jr., the \tev sources were fit with a Gaussian
    spatial model so the plotted \gev and \tev extensions were first
    converted to \rsixeight (see section~\ref{compare_source_size}).
    Because of their large size, the RX J1713.7-2946 and Vela Jr.
    were not directly fit in \tev and do not have errors associated with
    their measurements. The \tev radius of RX J1713.7-2946 is taken to
    be 0.5\deg and Vela Jr. is taken to be 1\deg. For these two sources,
    the LAT size is not converted to \rsixeight but the disk extension
    is directly compared.  This plot shows that there is good correlation
    between the \gev and \tev sizes of sources.
    }\label{gev_vs_tev_plot}
  \end{figure}

\clearpage
\begin{figure}
  \begin{center}
    \includegraphics{summary_plots/gev_vs_tev_histogram.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/extension_histogram/v2/extension_histogram.py
    \end{center}
    \caption{
    A comparison of the size of detected LAT and H.E.S.S sources.
    The extension of the 22 extended LAT sources comes from
    table~\ref{known_extended_sources}, table~\ref{new_ext_srcs_table},
    and the LAT extension of Vela X is taken from (\cite{velax}). The
    Centarus A lobes, which have an extension $~10\deg$, are not
    included in the plot (\cite{cen_a_lat}).  The \tev extension of
    the 42 extended H.E.S.S. sources comes from the H.E.S.S. Source
    Catalog (\cite{hesscat}).  Except for RX J1713.7-2946 and Vela Jr.,
    the H.E.S.S. sources were fit with a Gaussian spatial model and
    so the LAT and H.E.S.S. sizes are first converted to \rsixeight.
    (see section~\ref{compare_source_size}). Beacuse the spatial
    morphology of RX J1713.7-2946 and Vela Jr. is poorly matched by a
    gaussian, the \tev sizes are assuming a uniform disk spatial model
    and so neither the \gev nor \tev sizes were converted.  This plot
    shows that \tev experiments are sensitive to smaller extended sources.
    }\label{gev_vs_tev_histogram}
  \end{figure}

>>>>

Using two years of all sky surveying data, the LAT Collaboration
published a catalog of 1873 \gev sources\cite{second_cat}. Many of these can
be positionally associated with a variety of source classes including
Active Galactic Nuclei (AGN), pulsars, galaxies, and supernova
remnants (SNRs).  While many of these classes can be spatially
resolved when observed at other frequencies, detection an extension at
GeV energies by the LAT is complicated by the rather large
point-spread function (PSF) of the instrument (the angular resolution
for events converting in the front part of the detector, defined as
the 68\% containment radius for a point source, is approximately
$3.5^{\circ}$ at 100 MeV, improving to about 0.1$^{\circ}$ at 10 GeV).

Several of these source classes are extended (and close) enough to
potentially be extended. In particular supernova remnants, pulsar wind
nebulae, molecular clouds, normal galaxies, galaxy clusters or dark
matter satellites can (in favorable cases) be expected to be extended
beyond the LAT's PSF. The LAT collaboration has previously reported on
the discovered of five spatially extended supernova remnants, IC443,
W28, W44, W51C, and RX J1713.7-3946
(\cite{ic443,w28,w44,w51c,rx_j1713_lat}). In addition, three extended
pulsar wind nebulae (PWNe) were detected: MSH 15-52, Vela~X, and
HESS\,J1825-137 (\cite{msh1552,velax,fermi_hess_j1825}), as well as
two close-by galaxies, the Large Magellanic and Small Magalenic Cloud
(\cite{lmc,smc}), and finally the radio galaxy Centarus A
(\cite{cen_a_lat}) .

In addition to these GeV $\gamma$-ray detections, a variety of
spatially extended sources have been detected in the Galactic plane at
\tev energies using Air Cherenkov detectors. Most prominent was a
survey of the Galactic plane using the High Energy Stereoscopic System
(H.E.S.S) which presented 14 spatially extended sources
(\cite{hess_plane_survey}) sources with extensions varying from
$~0.1\deg$ to $~0.25\deg$. In fact, within our Galaxy only very few
sources (most notably the $\gamma$-ray binaries
LS\,5039\cite{HESSLS5039}, LS I+61-303\cite{MAGICLSI, VERITASLSI} and
HESS\,J0632+057\cite{HESS0632} are amongst the few Galactic sources
for which no significant extension was detected at \tev
energies. Given these findings, it can be expected that if the source
population at GeV energies has similar properties, that at least some
of the Fermi-LAT sources in our Galaxy should be spatially extended at
GeV energies.

Being able to spatially resolve the \gev emission from a source is
important for several reasons. Because of the large PSF and the large
candidate source density in our Galaxy, source confusion is a severe
problem for the Fermi-LAT. Finding a matching morphology can be used
to uniquely identify the source.  Because of the strongly varying PSF
of the LAT, the spatial and spectral information about a source do not
simply decouple. Therefore, a biased spatial model will bias the
spectral model of the source and studying the extension of the source
can correct for this.  Furthermore, correctly modeling a source's
extension is important for improving the overall model of the sky and
removing excess residuals.  These residuals manifested themselves in
1FGL as spurious point sources in the region region surrounding the
Large Magellanic Cloud (\cite{first_cat}).  For this reason, we are
motivated to take the 1873 sources that were detected in 2FGL and
apply a search for those sources which are spatially extended.

@@

@@

This source is spatially coincident with the extended \tev source HESS
J1616-508 (\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1613.6-5100c},
contours of HESS J1616-508 are overlaid on 2FGL J1615.0-5051.
The H.E.S.S. experiment measured an extension $0.136\pm 0.008$
when fitting the source with an elliptical Gaussian spatial model.
This size corresponds to a 68\% source containment radius of
$\rsixeight\sim0.21\deg\pm0.01$. This size is comparable to the LAT size
$\rsixeight=0.27\deg\pm0.03$ (see section~\ref{compare_source_size}).
Figure~\ref{hess_seds} shows that the LAT spectrum of
2FGL J1615.0-5051 connects smoothly to the 
H.E.S.S spectrum of HESS J1616-508.

@@

This source is spatially coincident with the extended \tev source HESS
J1614-518 (\cite{hess_plane_survey}). In figure~\ref{1FGL_J1613.6-5100c},
contours of HESS J1614-518 are overlaid on 2FGL J1615.2-5138.
The H.E.S.S. experiment measured a Gaussian extension of
$\sigma=0.23\deg\pm0.02\deg$ and $\sigma=0.15\pm0.02$ in the semi-major
and semi-minor axis. This size corresponds to a 68\% containment size
of $\rsixeight=0.35\pm0.03\deg$ and $0.23\pm0.03$.  This elliptical size
matches well the LAT size of $\rsixeight\sim0.34\pm0.02$.
Figure~\ref{hess_seds} shows that the LAT spectrum of
2FGL J1615.2-5138 connects smoothly to the 
H.E.S.S spectrum of HESS J1614-518.

@@

This extended source is spatially coincident with the extended \tev source HESS
J1632-478 (\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1632.9-4802c},
contours of  HESS J1632-478 are overlaid on 2FGL J1635.4-4717c.
H.E.S.S measured an extension of $\sigma=0.21\pm0.05$ and
$0.06\pm0.04$ along the semi-major and semi-minor axes when fitting the source
with an elliptical Gaussian spatial model.  This corresponds to a 68\%
containment size of $\rsixeight=0.31\pm0.08$ and $0.09\pm0.06$ along the
semi-major and semi-minor axis. This size is consistent with the LAT size of
$\rsixeight=0.36\pm0.03$.
Figure~\ref{hess_seds} shows that the LAT spectrum of
2FGL J1635.4-4717c connects smoothly to the 
H.E.S.S spectrum of HESS J1632-478.

@@

This source is spatially coincident with the \tev source HESS J1837-069
(\cite{hess_plane_survey}).  In figure~\ref{1FGL_J1837.5-0659c},
contours of are H.E.S.S J1837-069 are overlaid on 2FGL
J1837.3-0700c. H.E.S.S. measured an extension of $\sigma=0.12\pm0.02$
and $0.05\pm0.02$ along the semi-major and semi-minor axis when fitting
the source with an elliptical Gaussian spatial model.  This corresponds to
a 68\% containment radius of $\rsixeight=0.18\pm0.03$ and $0.08\pm0.03$
along the semi-major and semi-minor axis. The size is comparable to
LAT which fit a 68\% containment radius of $\rsixeight=0.29\pm0.07$.
The LAT size is larger than the H.E.S.S. size but the sizes are almost
almost within the $1\sigma$ errors. It is possible that the morphologies
are different but this is not very significant.  A possible reason is
confusion could be that the LAT source may be partially fitting other
sources in the region.  Figure~\ref{hess_seds} shows that the LAT
spectrum of 2FGL J1837.3-0700c connects smoothly to the H.E.S.S spectrum
of HESS J1837-069.

@@

This source is spatially conincident with the SNR RX J1713.7-3946 and was
recently reported by (\cite{rx_j1713_lat}).  To analyze this source,
we used the same background model as the recent LAT publication.
2FGL J1715.4-4024c is spatially coincident with Source A and was
moved to the published location of $(ra,dec)=(258.84,-40.46)$. Source
B was added at $(ra,dec)=(258.71,-38.70)$ and Source C was added at
$(ra,dec)=(257.47,-39.75)$.  Figure \ref{2FGL_J1712.4-3941} shows
the position of these nearby sources on top of a contour of SNR RX
J1713.7-3946 seen in \tev by H.E.S.S (\cite{rx_j1713_hess}).

@@

\begin{figure}
  \begin{center}
    \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_RX_J1713.7-3946}
    % this plot came from
  % /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/source_plots/RX_J1713.7-3946/v6/run.py
  \end{center}
  \caption{A Galactic and isotropic diffuse subtracted 1 \gev to 100
  \gev counts map of the region around 2FGL J1712.4-3941 smoothed by a
  0.25\deg Gaussian kernel.  This source is spatially coincident with
  RX J1713.7-3946 was was recently reported by (\cite{rx_j1713_lat}).
  The light blue contours are the \tev emission reported by
  H.E.S.S. (\cite{rx_j1713_hess}).  The region was analyzed with the
  same background model as (\cite{rx_j1713_lat}).  Source A (blue cross)
  is spatially coincident with 2FGL J1715.4-4024c (blue star) so 2FGL
  J1715.4-4024c was moved to the position of source A.  The grey crosses
  represent from left to right the position of source B and C which were
  added to the background model. Plot (a) has the diffuse emisission model
  subtracted and plot (b) has the diffuse emission and all background
  sources also subtracted.
  }\label{2FGL_J1712.4-3941}
\end{figure}

@@

\clearpage
\begin{figure}
  \begin{center}
    \includegraphics{summary_plots/hess_seds.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/seds/v2/plot_seds/plot_seds.py
    \end{center}
    \caption{
    }\label{hess_seds}
  \end{figure}

@@

\clearpage
\begin{figure}
  \begin{center}
    \includegraphics{summary_plots/gev_vs_tev_plot.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/compare_tev_gev_sizes/v1/run.py
    \end{center}
    \caption{
    A comparison of the \gev and \tev sizes of LAT
    extended sources detected by \tev instruments.  The \tev
    extensions of W30, 2FGL J1837.3-0700c, 2FGL J1632.4-4753c,
    2FGL J1615.0-5051, and 2FGL J1615.2-5138 are from the
    H.E.S.S. Galactic plane survey (\cite{hess_plane_survey}).
    The \tev extensions of MSH 15-52, HESS J1825-137, Vela X,
    Vela Jr., RX J1713.7-3946 and W28 are respectively from H.E.S.S.
    (\cite{msh_15_52_hess,hess_j1825_hess,vela_x_hess,vela_jr_hess,rx_j1713_hess,w28_hess}).
    The \tev extension of IC443 is from Veritas (\cite{ic443_veritas})
    and W51c is from Magic (\cite{w51c_with_magic_at_fermi_symposium}).
    The LAT extension of Vela X is from (\cite{velax}).  Except for RX
    J1713.7-3946 Vela Jr., the \tev sources were fit with a Gaussian
    spatial model so the plotted \gev and \tev extensions were first
    converted to \rsixeight (see section~\ref{compare_source_size}).
    Because of their large size, the RX J1713.7-3946 and Vela Jr.
    were not directly fit in \tev and do not have errors associated with
    their measurements. The \tev radius of RX J1713.7-3946 is taken to
    be 0.5\deg and Vela Jr. is taken to be 1\deg. For these two sources,
    the LAT size is not converted to \rsixeight but the disk extension
    is directly compared.  This plot shows that there is good correlation
    between the \gev and \tev sizes of sources.
    }\label{gev_vs_tev_plot}
  \end{figure}

\clearpage
\begin{figure}
  \begin{center}
    \includegraphics{summary_plots/gev_vs_tev_histogram.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/2FGL/plots_for_paper/extension_histogram/v2/extension_histogram.py
    \end{center}
    \caption{
    A comparison of the size of detected LAT and H.E.S.S sources.
    The extension of the 22 extended LAT sources comes from
    table~\ref{known_extended_sources}, table~\ref{new_ext_srcs_table},
    and the LAT extension of Vela X is taken from (\cite{velax}). The
    Centarus A lobes, which have an extension $~10\deg$, are not
    included in the plot (\cite{cen_a_lat}).  The \tev extension of
    the 42 extended H.E.S.S. sources comes from the H.E.S.S. Source
    Catalog (\cite{hesscat}).  Except for RX J1713.7-3946 and Vela Jr.,
    the H.E.S.S. sources were fit with a Gaussian spatial model and
    so the LAT and H.E.S.S. sizes are first converted to \rsixeight.
    (see section~\ref{compare_source_size}). Beacuse the spatial
    morphology of RX J1713.7-3946 and Vela Jr. is poorly matched by a
    gaussian, the \tev sizes are assuming a uniform disk spatial model
    and so neither the \gev nor \tev sizes were converted.  This plot
    shows that \tev experiments are sensitive to smaller extended sources.
    }\label{gev_vs_tev_histogram}
  \end{figure}
